\titletext{}

### 2021-04-01 manuscript/regional_befr_manuscript.tex
Edits following comments

\begin{itemize}
	\item{Positive biodiversity-ecosystem function relationships (BEFRs) have been widely documented, but it is unclear if BEFRs should be expected in disturbance-driven systems. Disturbance may limit competition and niche differentiation, which are frequently posited to underlie BEFRs. We provide the first exploration of the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density and vegetation type.}
	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity. The observed BEFR was highly dependent on organismal density, with a minimum threshold of c. 180 tree stems ha\textsuperscript{-1}. We found that water availability mainly affects biomass indirectly, via increasing species diversity.}
	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed savannas and woodlands. We suggest that tree diversity is an under-appreciated determinant of wooded ecosystem structure and function.}
\end{itemize}

@@

% BEFR is important to study, setting the scene - DONE
Understanding the relationship between biodiversity and ecosystem function has become a central endeavour in ecological science, as we seek to predict the consequences of global biodiversity change \citep{Naeem2012}. Over the past two decades, study of the Biodiversity-Ecosystem Function Relationship (BEFR) has grown from small-scale experimental studies mostly in temperate grasslands \citep{Cardinale2009, Tilman1994, Tilman2014}, to observational studies in natural ecosystems \citep{Plas2019}. While positive BEFRs which align with theory have been frequently reported, a complex picture has emerged whereby the strength and direction of the BEFR varies depending on the ecosystem studied \citep{Liang2016}, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}. The goal now should be to study the BEFR in different environmental and ecological contexts, in order to develop an ecosystem-agnostic understanding of the complex interactions between biodiversity, abiotic environment, and ecosystem function.

% Ecosystem function definition and mechanisms of the BEFR - DONE
Ecosystem functions are defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}. Woody productivity and biomass storage are two of the most commonly studied ecosystem functions in forests and other wooded ecosystems (e.g. savannas) \citep{Brockerhoff2017}, due to their importance in the global carbon cycle \citep{Pan2011} and their potential leverage as a tool to mitigate the effects of anthropogenic climate change while maintaining biodiversity \citep{Pichancourt2013}. Theory predicts that biodiversity will have a positive effect on ecosystem function via three principal mechanisms: 1) niche complementarity, whereby niche partitioning and ecosystem resource use efficiency is increased in diverse communities, minimising negative density dependent effects such as conspecific competition and abundance of species-specific natural enemies \citep{Barry2019}; 2) selection effects, whereby diverse communities are more likely to include a high-yield species; and 3) facilitation effects, whereby diverse communities are more likely to contain species combinations which enhance each others' functional contribution \citep{Wright2017}.

@@

A number of studies in miombo woodlands, the dominant savanna type in southern Africa \citep{Campbell1996}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. These studies however, lacked the spatial extent required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR. Studies of the BEFR often find that within a small spatial extent (<2500 km\textsuperscript{2}), biodiversity shows a strong effect on ecosystem function, but at broader extents (>100,000s km\textsuperscript{2}) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Gonzalez2020}. In West Africa, \citet{Mensah2020} found that woodlands and forests showed a positive effect of tree species richness on above-ground carbon, while sparse savannas did not, implying that tree stem density and the presence of dominant large trees may affect the strength of the observed BEFR. In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity via its effect on functional trait values. In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency. Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, allowing weak competitors to co-exist where otherwise they would be excluded \citep{Mason2011}. While these studies together offer some glimpse into how the BEFR may operate in disturbance-prone systems, none provide a full and comprehensive assessment of the interacting effects of diversity, abiotic environment, disturbance and ecosystem function, particularly for wooded ecosystems. Additionally, due to differences in community assembly, evolutionary history and contemporary drivers between the neotropics and the African tropics, inferences from one continent cannot necessarily be applied to the other \citep{Dexter2015}.

@@

% Situation - DONE
The study used \nplots{} woodland monitoring plots from a larger pool of 5395 plots in the larger SEOSAW database \citep{SEOSAW2020}, located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from c. 460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to c. 1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

% Plots chosen - DONE
Plots were chosen from the SEOSAW database based on the quality and completeness of data collection, and plot configuration. Plot vegetation was identified under the broad term of `savanna', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands, or savannas \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate biomass estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered here a separate biome with very different species composition \citep{Parr2014}. \nzam{} plots within the SEOSAW database were arranged in clusters of four 20x50 m plots, with 20 m between plots. Plots within each spatial cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.

@@

In the sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}e). There appears to be a minimum stem density threshold at c. 180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass (\autoref{sem_struc_stems_ha}b). The effect of structural diversity on AGB appears to remain constant with increasing stem density (\autoref{sem_struc_stems_ha}d). The indirect effect of tree species diversity on AGB via structural diversity increases as stem density increases (\autoref{sem_struc_stems_ha}c).

@@

As time since disturbance increases in any wooded ecosystem, it is expected that both species diversity and above-ground biomass will increase until an equilibrium is reached or another disturbance event occurs \citep{Huston2014}. Unless this effect is accounted for, there is the potential for mistaken causality as both diversity and biomass will correlate. In our model, time since disturbance is accounted for within each plot via the stem density term. Disturbance reduces stem density, which is expected to increase until the effects of competition preclude further increase \citep{Johnson2012}. Furthermore, our rarefied measure of species diversity accounts for variation in sampling effort and is therefore independent of stem density. Tree species richness should also increase with time since disturbance as with increased stem density the likelihood of including a new species also increases. Outside of the stem density effect, there are multiple causes for variation in tree species diversity in this study. Vegetation types and localities differ in their available species pool, for example. Variation in abiotic environmental factors will also affect species accumulation.

@@

A negative total effect of soil fertility on AGB is in contrast to other studies in the region and general ecological theory, which predicts a positive effect of soil nutrients on biomass \citep{Scarascia2000}. The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density. The direct effect on AGB however, remained positive and marginally significant, as expected. Model estimates of the effect of soil on AGB were poorly constrained compared with other latent variables. This wide standard error on the model predictions is possibly due to the coarseness and nature of the soil data we used. SoilGrids provides modelled data at 250 m resolution, while sostructure and nutrient content varies at much finer scales in southern African woodlands \citep{Muledi2017, Bucini2007}, notably by termite mounds which offer small areas of highly fertile soil with peculiar tree species diversity \citep{Joseph2014}. It is therefore not surprising that this model path is poorly constrained. \citet{Lehmann2014} found similarly weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, and fire to predict tree basal area in southern African woodlands. Plot-specific soil data are time-consuming to collect and difficult to compare across studies when different protocols are used. Our study points to the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands.

@@

\begin{landscape}
\section{Figures}
\begin{figure}[H]
\centering
	\includegraphics[width=1.4\textwidth]{clust_map}
	\caption{The locations of the \nplots{} plots used in this study, with respect to the distribution of miombo woodland vegetation according to \citet{White1987}. Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
	\label{clust_map}
\end{figure}
\end{landscape}
\newpage{}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{concept.drawio}
	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stem density, and AGB. Hypothesised paths of causation are depicted as arrows from predictor to response.}
	\label{con_mod}
\end{figure}
\newpage{}

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
	\caption{Line plots showing the variation in SEM path coefficients across datasets with different mean stem density. Smoothed lines are loess curves with $\pm1$ standard error shaded bars. AGB = Above-Ground woody Biomass, arrows in plot titles indicate causal paths in SEM models.}
	\label{sem_struc_stems_ha}
\end{figure}
\newpage{}

>>>>

\begin{itemize}
	\item{Positive biodiversity-ecosystem function relationships (BEFRs) have been widely documented, but it is unclear if BEFRs should be expected in disturbance-driven systems. Disturbance may limit competition and niche differentiation, which are frequently posited to underlie BEFRs. We provide the first exploration of the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density and vegetation type.}
	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity. The observed BEFR was highly dependent on organismal density, with a minimum threshold of c. 180 mature stems ha\textsuperscript{-1}. We found that water availability mainly affects biomass indirectly, via increasing species diversity.}
	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed savannas and woodlands. We suggest that tree diversity is an under-appreciated determinant of wooded ecosystem structure and function.}
\end{itemize}

@@

Understanding the relationship between biodiversity and ecosystem function has become a central endeavour in ecological science, as we seek to predict the consequences of global biodiversity change \citep{Naeem2012}. Over the past two decades, study of the Biodiversity-Ecosystem Function Relationship (BEFR) has grown from small-scale experimental studies mostly in temperate grasslands \citep{Cardinale2009, Tilman1994, Tilman2014}, to observational studies in natural ecosystems \citep{Plas2019}. While positive BEFRs which align with theory have been frequently reported, a complex picture has emerged whereby the strength and direction of the BEFR varies depending on the ecosystem studied \citep{Liang2016}, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}. The goal now should be to study the BEFR in different environmental and ecological contexts, in order to develop an ecosystem-agnostic understanding of the complex interactions between biodiversity, abiotic environment, and ecosystem function.

Ecosystem functions are defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}. Woody productivity and biomass storage are two of the most commonly studied ecosystem functions in forests and other wooded ecosystems (e.g. savannas) \citep{Brockerhoff2017}, due to their importance in the global carbon cycle \citep{Pan2011} and their potential leverage as a tool to mitigate the effects of anthropogenic climate change while maintaining biodiversity \citep{Pichancourt2013}. Theory predicts that biodiversity will have a positive effect on ecosystem function via three principal mechanisms: 1) niche complementarity, whereby niche partitioning and ecosystem resource use efficiency is increased in diverse communities, minimising negative density dependent effects such as conspecific competition and abundance of species-specific natural enemies \citep{Barry2019}; 2) selection effects, whereby diverse communities are more likely to include a high-yield species; and 3) facilitation effects, whereby diverse communities are more likely to contain species combinations which enhance each others' functional contribution \citep{Wright2017}.

@@

A number of studies in miombo woodlands, the dominant savanna type in southern Africa \citep{Campbell1996}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. These studies however, lacked the spatial extent required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR. Studies of the BEFR often find that within a small spatial extent (<2500 km\textsuperscript{2}), biodiversity shows a strong effect on ecosystem function, but at broader extents (>100,000s km\textsuperscript{2}) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Gonzalez2020}. In West Africa, \citet{Mensah2020} found that woodlands and forests showed a positive effect of tree species richness on above-ground carbon, while sparse savannas did not, implying that tree stem density and the presence of dominant large trees may affect the strength of the observed BEFR. In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity via its effect on functional trait values. In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency. Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, allowing weak competitors to co-exist where otherwise they would be excluded \citep{Mason2011}, thus weakening the observable BEFR. Thus, there is a case that in highly disturbed systems, a relationship between diversity and ecosystem function may not exist at all. While these studies together offer some glimpse into how the BEFR may operate in disturbance-prone systems, none provide a full and comprehensive assessment of the interacting effects of diversity, abiotic environment, disturbance and ecosystem function, particularly for wooded ecosystems. Additionally, due to differences in community assembly, evolutionary history and contemporary drivers between the neotropics and the African tropics, inferences from one continent cannot necessarily be applied to the other \citep{Dexter2015}.

@@

% Situation - DONE
The study used \nplots{} woodland monitoring plots from a larger pool of 5395 plots in the larger SEOSAW database \citep{SEOSAW2020}, located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from c. 460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to c. 1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temn intermediate erature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

% Plots chosen - DONE
Plots were chosen from the SEOSAW database based on the quality and completeness of data collection, and plot configuration. Plot vegetation was identified under the broad term of `savanna', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands, or savannas \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction, experimental treatments such as prescribed burning or herbivore exclusion, or containing termite mounds were excluded from the initial pool. Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate biomass estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered here a separate biome with very different species composition \citep{Parr2014}. \nzam{} plots within the SEOSAW database were arranged in clusters of four 20x50 m plots, with 20 m between plots. Plots within each spatial cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.

@@

In the sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}e). There appears to be a minimum stem density threshold at c. 180 stems >10 cm DBH ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass (\autoref{sem_struc_stems_ha}b). The effect of structural diversity on AGB appears to remain constant with increasing stem density (\autoref{sem_struc_stems_ha}d). The indirect effect of tree species diversity on AGB via structural diversity increases as stem density increases (\autoref{sem_struc_stems_ha}c).

@@

As time since disturbance increases in any wooded ecosystem, it is expected that both species diversity and above-ground biomass will increase until an equilibrium is reached or another disturbance event occurs \citep{Huston2014}. Unless this effect is accounted for, there is the potential for mistaken causality as both diversity and biomass will correlate. In our model, time since disturbance is accounted for within each plot via the stem density term. Disturbance reduces stem density of large stems (>10 cm DBH), which is expected to increase until the effects of competition preclude further increase \citep{Johnson2012}. Furthermore, our rarefied measure of species diversity accounts for variation in sampling effort and is therefore independent of stem density. Tree species richness should also increase with time since disturbance as with increased stem density the likelihood of including a new species also increases. Outside of the stem density effect, there are multiple causes for variation in tree species diversity in this study. Vegetation types and localities differ in their available species pool, for example. Variation in abiotic environmental factors will also affect species accumulation.

@@

A negative total effect of soil fertility on AGB is in contrast to other studies in the region and general ecological theory, which predicts a positive effect of soil nutrients on biomass \citep{Scarascia2000}. The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density. The direct effect on AGB however, remained positive and marginally significant, as expected. Model estimates of the effect of soil on AGB were poorly constrained compared with other latent variables. This wide standard error on the model predictions is possibly due to the coarseness and nature of the soil data we used. SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales in southern African woodlands \citep{Muledi2017, Bucini2007}. It is therefore not surprising that this model path is poorly constrained. \citet{Lehmann2014} found similarly weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, and fire to predict tree basal area in southern African woodlands. Plot-specific soil data are time-consuming to collect and difficult to compare across studies when different protocols are used. Our study points to the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands.

@@

\begin{landscape}
\section{Figures}
\begin{figure}[H]
\centering
	\includegraphics[width=1.4\textwidth]{clust_map}
	\caption{The locations of the \nplots{} plots used in this study, with respect to the distribution of mesic savanna vegetation according to \citet{White1987}. Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
	\label{clust_map}
\end{figure}
\end{landscape}
\newpage{}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{concept.drawio}
	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stem density, and AGB. Hypothesised paths of causation are depicted as arrows from predictor to response. Open arrow heads track the direction of each arrow along its path.}
	\label{con_mod}
\end{figure}
\newpage{}

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
	\caption{Line plots showing the variation in SEM path coefficients across datasets with different mean stem density. Smoothed lines are loess curves with $\pm1$ standard error shaded bars. AGB = Above-Ground woody Biomass, arrows in plot titles indicate causal paths in SEM models. Where multiple arrows are present, as in c), this indicates an indirect pathway via an intermediate variable.}
	\label{sem_struc_stems_ha}
\end{figure}
\newpage{}
